Sparse Function-space Representation \\ of Neural Networks

### 2023-05-23 paper/main.tex
updated image stuff

\subsection{Supervised learning on image data sets}
\label{sec:image}
Similarly to the UCI experiments, we seek to demonstrate \our on image data that features more complicated data manifolds and model structures.
We consider the MNIST and Fashion-MNIST (FMNIST) classification tasks, and use an MLP and a CNN architecture respectively (matching the setup in \cite{immer2021improving}).
After training, we computed the duals for \our and use the model for prediction (full experiment details in \cref{app:image}).
In \cref{tbl:imagesuper}, we compare to baselines also used in \cite{immer2021improving}.
We report accuracy (ACC), NLPD, and expected calibration error (ECE).
For \our and the GP predictive subset method by \cite{immer2021improving}, we provide results when predicting with GP mean and when using the neural network (NN) and
fix the number of inducing points $M=1024$.
For completeness, we also include the GP predictive result from \cite{immer2021improving} which used more inducing points $M=3200$.
The results in \cref{tbl:imagesuper} agree with the conclusions drawn from the UCI experiments showing that \our is able to more efficiently capture the posterior using a sparse set of points.
They further demonstrate that \our's predictive mean (GP) significantly outperforms the GP predictive subset.
This is an interesting result as we cannot rely on the NN's prediction if we wish to condition on new data, using the dual conditioning from \cref{sec:sequential}.
In this setting, we must rely on our GP's predictive mean.
% Other approaches, \cite{immer2021improving} never inten
% The resutls demonstrate that \our's predictive GP mean significantly outperforms the
These results on image data sets motivate experiments beyond the supervised learning setting, which we now detail.

@@

\begin{table}[!ht]
\centering
\begin{tabular}{@{}ccc@{}}
\toprule
Loss        & der 1 & der 2 \\ \midrule
binomial    &       &       \\
mse         &       &       \\
categorical &       &       \\ \bottomrule
\end{tabular}
\end{table}

\section{\our for continual learning}
\label{app:cl}
\subsection{Derivation of the regularizer term}
For continual learning, the model cannot access the training data from previous tasks, denoted here as $\dataset_\textrm{old} = \bigcup_{s=1}^{t-1} \dataset_s$, and for each new tasks receives a new chunk of data $\dataset_\textrm{new} = \dataset_t$. On the other hand, in the supervised learning setting (described in \cref{sec:methods}) the neural network model has access to all the training data at once, \ie, the objective we are trying to optimize is
\begin{equation}
	\weights^* = \argmin_\weights \mathcal{L}(\dataset_\textrm{old} \cup \dataset_\textrm{new}, \weights) = \argmin_\weights \sum_{i \in \dataset_\textrm{old} \cup \dataset_\textrm{new}}\ell(f_\weights(\mathbf{x}_{i}), y_i) + \delta \mathcal{R}(\weights).
\end{equation} 
Our goal in the continual setting is to replace the missing data coming from $\dataset_\textrm{old}$ by resorting to a functional regularizer that accounts for that by forcing the predictions of the neural networks over the inducing points $\mathcal{Z}_t$ to come with high probability from $f_\weights(\vz_i) \sim q(\vu),  \forall \vz_i \in \mathcal{Z}_t$. As shown in \citep{khan2017conjugate, adam2021dual}, we can view the dual parameters derived in \cref{sec:sparse-dual-gp}, as {\color{red} TODO}.  Then, we can rewrite the sparse GP posterior for $q(\vu)$ in terms of the following approximate normal distributions:
%
\begin{align}
	q(\vu)
	& \propto p_\weights(\vu) \Norm(\tilde{\vy}; \vu, \MB_\vu)\\
	& \propto  \Norm(0, \MKzz) \prod_{i \in \dataset_\textrm{old}} \exp \! \left(-\frac{\hat{\beta}_i}{2}(\tilde{y}_i - \va_i^\top \vu)^2 \right) \\
	& \propto \Norm(0, \MKzz) \, \exp\left((\tilde{\vy} - \vu)^\top \bar{\MB}^{-1}(\tilde{\vy} - \vu)\right)
\end{align}
% The first normals are the approximate likelihoods 
where $\va_i^\top = \vkzi^\top \MKzz^{-1}$ and $\tilde{y}_i = \hat{\alpha}_i + f_{\vw^*}(\vx_i)\hat{\beta}_i$.
The sparse GP posterior $q(\vu)$ is proportional to the product between the prior over the weights $p_\weights(\vu)$ and a Gaussian term $\Norm(\tilde{\vy}; \vu, \MB_\vu)$, which arises from {\color{red} TODO}

Replacing $\va_i^\top \vu$ with $\vx_i^{\top}\vw$, we can see the similarity between the sparse GP and a linear model where the kernel matrix determines the prior. To arrive at the MVN Gaussian, we need to rearrange the quadratic terms to match the prior sufficient statistics. Performing conjugate summation of prior and likelihood, we would arrive at the sparse GP posterior at the inducing points. The correlated covariance structure $\MSigma$ and the corresponding $\tilde{\vy}$ are simply a different form of the sparse dual variables,
\begin{equation}
\quad \bar{\MB}^{-1} =  \MKzz \MB_\vu^{-1} \MKzz \quad \tilde{\vy} = \MKzz \MB_{\vu}\valpha_{\vu}.
\end{equation}
Written as above, we can see that the spare dual variables can be interpreted as a sparse MVN representation.

\subsection{Extending the CL regularizer to multi-class settings} 
\label{sec:cl_multioutput}
In the multi-class setting, the neural network can be seen as a function approximator that maps inputs $\vx \in \inputDomain$ to  the logits $\textit{\textbf{f}} \in \outputDomain$, \ie, $f_\mathbf{w}: \inputDomain \to \outputDomain$. As a consequence, the Jacobian matrix is $\Jac{\weights}{\vx} \coloneqq \left[ \nabla_\weights f_\weights(\vx)\right]^\top \in \R^{C \times P}$ and the kernel of the associated sparse GP derived in \cref{sec:sfr}, $\kappa(\vx, \vx')$ is a tensor in $\R^{C \times P \times P}$. This means that we need compute the dual parameters and the matrix $\bar{\MB}^{-1}_{t, c}$ for all the function's outputs. If we denote $\kappa_c(\vx, \vx') \in \R$, as the kernel related to the $c$\textsuperscript{th} output function, \ie, the logit for class $c$, we can write the dual parameters as
%
\begin{equation} \textstyle
  \valpha_{\vu, c}  =  \sum_{i=1}^N  \vkzic \, \hat{\alpha}_{i, c} \in \R^{M}
  \quad \text{and} \quad
  \MBeta_{\vu, c} =  \sum_{i=1}^N \vkzic \,\hat{\beta}_{i, c} \, \vkzic^{\T} \in \R^{M \times M} ,    
\label{eq:dual_sparse}
\end{equation}
%
where $\vkzic$ is the kernel $\kappa_c(\cdot, \cdot)$ computed for the inducing points $\vz \in \mathcal{Z}$ and the input point $\vx_i$ for class $c$. In this setting, the dual variables are
\begin{equation}
\begin{aligned}
  \hat{\alpha}_{i,c} &\coloneqq  \left[\hat{\valpha}_{i}\right]_c = \left[\nabla_{f}\log p(y_i \mid f) |_{f=f_i} \right]_c \in \R\\
  \hat{\beta}_{i,c} &\coloneqq \left[\textrm{diag}(\hat{\vbeta}_i)\right]_c = \left[ - \nabla^2_{ff}\log p(y_i \mid f) |_{f=f_i} \right]_c \in \R 
\end{aligned}
\end{equation}
where the operator $[ \cdot ]_c$ is taking the $c$\textsuperscript{th} element of a vector. In fact, in the multi-class setting $\hat{\valpha}_{i}$ is a vector in $\R^C$ and $\hat{\vbeta}_{i}$ a matrix in $\R^{C \times C}$.
For the \our-regularizer, at the end of each task we compute the matrix $\bar{\MB}^{-1}_{t, c}$ as 
\begin{equation}\textstyle
 	\bar{\MB}^{-1}_{t, c} = \MKzzc^{-1} \MBeta_{\vu, c} \MKzzc^{-1} \in \R^{M \times M}, 
 	\quad \text{s.t.} \quad
 	\MBeta_{\vu, c} =  \sum_{i \in \dataset_t} \vkzi \,\hat{\beta}_{i, c} \, \vkzi^{\T}.
\end{equation}
In practice computing the kernel $\kappa_c$ for classes $c$, for which the model has not seen any data yet is not meaningful \todo{probably better to word this differently}, so we define $\bar{\MB}^{-1}_{t, c}$ to be  $\bar{\MB}^{-1}_{t, c} = \MI_M$ $\forall c \not\in \mathcal{C}_t$, where $\mathcal{C}_t$ is the set of classes observed up to task t. For example, for S-MNIST  we will have after task 1 $\mathcal{C}_1 = \{0, 1\}$, then after task 2 $\mathcal{C}_1 = \{0, 1, 2, 3\}$, and so on. 
Finally, we can rewrite $\mathcal{R}_\mathrm{\our}$ as follows
\begin{equation}\label{eq-cl-regularizer}\textstyle
\begin{aligned}
  \mathcal{R}_\mathrm{\our}(\weights, \mathcal{M}) = \frac{1}{2} \sum_{s=1}^{t-1} \sum_{c=1}^{C} \frac{1}{M} 
	\left\lVert 
	\vu_c - \vu_{s,c} % f_{\weights_{s}}(\MZ_s)
	\right\rVert_{\bar{\MB}^{-1}_{s}}, \\
	\text{~with~} \vu_c \in \R^M \text{~with~} u_{i, c} = \left[ f_{\weights}(\vz_{i}) \right]_c,~\forall {\vz_i} \in \mathcal{Z}_s .
\end{aligned}
\end{equation}

@@

All neural network designs and related hyperparameters are set similarly to \citet{immer2021improving} for a fair comparison, for instance we train the $2$-layer MLPs with learning rate $0.001$ for $10000$ epochs, using the full dataset instead of batching. In addition, we vary the subset size $M \in  \{16, 32, 64, 128, 256, 512 \}$ to observe the impact of sparseness on the predictive performance, see \cref{tbl:uci-appendix} and \cref{tbl:uci-appendix-acc} for the complete NLPD and accuracy results. Our code implementation of the UCI experiments has been built on top of the code provided for \citet{immer2021improving}, using the \our methods provided in the supplementary material.

\begin{table}[t!] 
  \centering\scriptsize
  \caption{Complements \cref{tbl:uci} in the main paper: Negative log predictive density (NLPD, lower better) for our \our method compared to a GP subset based method~\cite{immer2021improving} which discards most of the data. These UCI data sets (num.\ of points $N$, input dim.\ $D$, num.\ classes $C$) demonstrate that even with a small number $M$ of inducing points our method is able to capture good predictive densities.} 
	\label{tbl:uci-appendix}
	% Control table spacing
	\renewcommand{\arraystretch}{1.}
	\setlength{\tabcolsep}{2pt}
	\setlength{\tblw}{0.25\textwidth}

@@

\begin{table}[t!] 
  \centering\scriptsize
  \caption{Complements \cref{tbl:uci-appendix}: We also report the accuracy for each number of inducing points $M$ for both our \our and the GP subset method. The differences in the accuracies is not as big as in NLPD which is not surprising as it is generally easier to capture first-order information with sparse methods. However, with small $M$ \our still shows clearly better performance.} 
	\label{tbl:uci-appendix-acc}
	% Control table spacing
	\renewcommand{\arraystretch}{1.}
	\setlength{\tabcolsep}{2pt}
	\setlength{\tblw}{0.25\textwidth}

@@

\subsection{Uncertainty benchmarks for image data}
\label{app:image}
Similarly to the supervised UCI benchmark tasks, a single image experiment consists of the following steps: first, we train a neural network (a MLP for MNIST, a CNN for FMNIST) in the classification task, using a prior precision $\delta$. After training, we prepare the \our method for predictions by computing the dual parameters, and then evaluate the validation set predictive performance of \our. While in principle our method's predictions consist of the GP mean and variance, we include results for using the neural network prediction as a mean instead. For comparison, we gather the GP subset method results for both the GP and neural network means. This experiment set-up is then replicated for a range of prior precision values and learning rates, as we perform a grid search over these hyperparameters by optimizing for the validation set NLPD. We run the experiment on $5$ seeds and pick the hyperparameter by the best mean performance over the seeds. The results reported in \cref{tbl:imagesuper} used a learning rate of $10^{-4}$, and a prior precision of $\delta = 0.0013$ for FMNIST and $\delta = 0.0001$ for MNIST \todo{double check}.

For the neural network training, we used the same set-up and neural network architectures as in \citet{immer2021improving}, then we trained them for $500$ epochs with batch size of $512$. In contrast to the GP subset implementation in \citet{immer2021improving}, we do not perform scaling of $\delta$ after the neural network has been trained, but use the same prior precision in \our as was used in the regularizer term while training. Avoiding this post-hoc tuning stage allows us to forgo the additional hyperparameter optimization step for the prior precision multiplier. We found the MNIST and FMNIST experiments to be numerically stable, even when not performing any clipping in computing the Hessian of the log likelihood, or adding jitter before Cholesky decomposition within the \our dual computation.

>>>>

\subsection{Supervised learning on image data sets}
\label{sec:image}
Similarly to the UCI experiments, we seek to demonstrate \our on image data that features more complicated data manifolds and model structures.
We consider the MNIST and Fashion-MNIST (FMNIST) classification tasks, and use an MLP and a CNN architecture respectively (matching the setup in \cite{immer2021improving}).
After training, we computed the duals for \our and use the model for prediction (full experiment details in \cref{app:image}).
In \cref{tbl:imagesuper}, we compare to baselines also used in \cite{immer2021improving}.
We report accuracy (ACC), NLPD, and expected calibration error (ECE).
For \our and the GP predictive subset method by \cite{immer2021improving}, we provide results when predicting with GP mean and when using the neural network (NN) and
fix the number of inducing points $M=1024$.
For completeness, we also include the GP predictive result from \cite{immer2021improving} which used more inducing points $M=3200$.
The results in \cref{tbl:imagesuper} agree with the conclusions drawn from the UCI experiments showing that \our is able to more efficiently capture the posterior using a sparse set of points.
They further demonstrate that \our's predictive mean (GP) significantly outperforms the GP predictive subset.
This is an interesting result as we cannot rely on the NN's prediction if we wish to condition on new data, using the dual conditioning from \label{sec:sequential}.
In this setting, we must rely on our GP's predictive mean.
% Other approaches, \cite{immer2021improving} never inten
% The resutls demonstrate that \our's predictive GP mean significantly outperforms the
These results on image data sets motivate experiments beyond the supervised learning setting, which we now detail.

@@

\section{\our for continual learning}
\label{app:cl}
\subsection{Derivation of the regularizer term}
An alternative view of the dual parameters is that they parameterize approximate likelihoods, as shown in \citep{khan2017conjugate, adam2021dual}. Then, we can rewrite the GP posterior for $q(\vu)$ in terms of the following approximate normal distributions:

\begin{equation}
\begin{aligned}
 q(\vu) 
 	&\propto \Norm(0, \MKzz) \prod_{i=1}^n \exp \! \left(-\frac{\tilde{\beta}_i}{2}(\tilde{y}_i - \va_i^\top \vu)^2 \right) \\
 	&= \Norm(0, \MKzz) \, \exp\left((\bar{\vy} - \vu)^\top \bar{\MSigma}^{-1}(\bar{\vy} - \vu)\right)  
\end{aligned}
\end{equation}

The first normals are the approximate likelihoods where $\va_i^\top = \vkzi^\top \MKzz^{-1}$ and $\tilde{y}_i = \hat{\alpha}_i + f_{\vw^*}(\vx_i)\hat{\beta}$,  replacing $\va_i^\top \vu$ with $\vx_i^{\top}\vw$ and we can see the similarity between the sparse GP and a linear model where the kernel matrix determines the prior. To arrive at the MVN Gaussian, we need to rearrange the quadratic terms to match the prior sufficient statistics we present a derivation in [appendix]. Performing conjugate summation of prior and likelihood, we would arrive at the sparse GP posterior at the inducing points. The correlated covariance structure $\MSigma$ and the corresponding $\tilde{\vy}$ are simply a different form of the sparse dual variables,
\begin{equation}
\quad \bar{\MSigma} =  \MKzz \vbeta_\vu^{-1} \MKzz \quad \bar{\vy} = \MKzz \vbeta_{\vu}\valpha_{\vu}.
\end{equation}
Written as above, we can see that the spare dual variables can be interpreted as a sparse MVN representation. We now show how we use this for a regularizer in continual learning with more details in [appendix].

\subsection{Extending the CL regularizer to multi-class settings} 
\label{sec:cl_multioutput}
In the multi-class setting, the neural network can be seen as a function approximator that maps inputs $\vx \in \inputDomain$ to  the logits $\textit{\textbf{f}} \in \outputDomain$, \ie, $f_\mathbf{w}: \inputDomain \to \outputDomain$. As a consequence, the Jacobian matrix is $\Jac{\weights}{\vx} \coloneqq \left[ \nabla_\weights f_\weights(\vx)\right]^\top \in \R^{C \times P}$ and the kernel of the associated sparse GP derived in \cref{sec:sfr}, $\kappa(\vx, \vx')$ is a tensor in $\R^{C \times P \times P}$. This means that we need compute the dual parameters and the matrix $\bar{\MB}^{-1}_{t, c}$ for all the function's outputs. If we denote $\kappa_c(\vx, \vx') \in \R$, as the kernel related to the $c$\textsuperscript{th} output function, \ie, the logit for class $c$, we can write the dual parameters as
%
\begin{equation} \textstyle
  \valpha_{\vu, c}  =  \sum_{i=1}^N  \vkzic \, \hat{\alpha}_{i, c} \in \R^{M}
  \quad \text{and} \quad
  \MBeta_{\vu, c} =  \sum_{i=1}^N \vkzic \,\hat{\beta}_{i, c} \, \vkzic^{\T} \in \R^{M \times M} ,    
\label{eq:dual_sparse}
\end{equation}
%
where $\vkzic$ is the kernel $\kappa_c(\cdot, \cdot)$ computed for the inducing points $\vz \in \mathcal{Z}$ and the input point $\vx_i$ for class $c$. In this setting, the dual variables are
\begin{equation}
\begin{aligned}
  \hat{\alpha}_{i,c} &\coloneqq  \left[\hat{\valpha}_{i}\right]_c = \left[\nabla_{f}\log p(y_i \mid f) |_{f=f_i} \right]_c \in \R\\
  \hat{\beta}_{i,c} &\coloneqq \left[\textrm{diag}(\hat{\vbeta}_i)\right]_c = \left[ - \nabla^2_{ff}\log p(y_i \mid f) |_{f=f_i} \right]_c \in \R 
\end{aligned}
\end{equation}
where the operator $[ \cdot ]_c$ is taking the $c$\textsuperscript{th} element of a vector. In fact, in the multi-class setting $\hat{\valpha}_{i}$ is a vector in $\R^C$ and $\hat{\vbeta}_{i}$ a matrix in $\R^{C \times C}$.
For the \our-regularizer, at the end of each task we compute the matrix $\bar{\MB}^{-1}_{t, c}$ as 
\begin{equation}\textstyle
 	\bar{\MB}^{-1}_{t, c} = \MKzzc^{-1} \MBeta_{\vu, c} \MKzzc^{-1} \in \R^{M \times M}, 
 	\quad \text{s.t.} \quad
 	\MBeta_{\vu, c} =  \sum_{i \in \dataset_t} \vkzi \,\hat{\beta}_{i, c} \, \vkzi^{\T}.
\end{equation}
However, in practice computing the kernel $\kappa_c$ for classes $c$, for which the model has not seen any data yet is not meaningful \todo{probably better to word this differently}, so we define $\bar{\MB}^{-1}_{t, c}$ to be  $\bar{\MB}^{-1}_{t, c} = \MI_M$ $\forall c \not\in \mathcal{C}_t$, where $\mathcal{C}_t$ is the set of classes observed up to task t. For example, for S-MNIST  we will have after task 1 $\mathcal{C}_1 = \{0, 1\}$, then after task 2 $\mathcal{C}_1 = \{0, 1, 2, 3\}$, and so on. 
Finally, we can rewrite $\mathcal{R}_\mathrm{\our}$ as follows
\begin{equation}\label{eq-cl-regularizer}\textstyle
\begin{aligned}
  \mathcal{R}_\mathrm{\our}(\weights, \mathcal{M}) = \frac{1}{2} \sum_{s=1}^{t-1} \sum_{c=1}^{C} \frac{1}{M} 
	\left\lVert 
	\vu_c - \vu_{s,c} % f_{\weights_{s}}(\MZ_s)
	\right\rVert_{\bar{\MB}^{-1}_{s}}, \\
	\text{~with~} \vu_c \in \R^M \text{~with~} u_{i, c} = \left[ f_{\weights}(\vz_{i}) \right]_c,~\forall {\vz_i} \in \mathcal{Z}_s .
\end{aligned}
\end{equation}

@@

\par All neural network designs and related hyperparameters are set similarly to \citet{immer2021improving} for a fair comparison, for instance we train the $2$-layer MLPs with learning rate $0.001$ for $10000$ epochs, using the full dataset instead of batching. In addition, we vary the subset size $M \in  \{16, 32, 64, 128, 256, 512 \}$ to observe the impact of sparseness on the predictive performance, see \cref{tbl:uci-appendix} and \cref{tbl:uci-appendix-acc} for the complete NLPD and accuracy results. Our code implementation of the UCI experiments has been built on top of the code provided for \citet{immer2021improving}, using the \our methods provided in the supplementary material.

\begin{table}[t!] 
  \centering\scriptsize
  \caption{Negative log predictive density (NLPD, lower better) for our \our method compared to a GP subset based method which discards most of the data. Even with a small $M$ our method is able to provide good results.} 
	\label{tbl:uci-appendix}
	% Control table spacing
	\renewcommand{\arraystretch}{1.}
	\setlength{\tabcolsep}{2pt}
	\setlength{\tblw}{0.15\textwidth}

@@

\begin{table}[t!] 
  \centering\scriptsize
  \caption{The accuracy of our \our method compared to a GP subset based method which discards most of the data. Even with a small $M$ our method is able to provide good results.} 
	\label{tbl:uci-appendix-acc}
	% Control table spacing
	\renewcommand{\arraystretch}{1.}
	\setlength{\tabcolsep}{2pt}
	\setlength{\tblw}{0.15\textwidth}

@@

\subsection{Uncertainty benchmarks for image data}
\label{app:image}
% Each experiment with the image data sets consisted of the following steps.
% First, we trained a neural network (MLP for MNIST, CNN for FMNIST) using a prior precision $\delta$.
% After training, we performed inference by calculating the dual parameters.
% Finally, we evaluated the predictive performance of \our on the validation set.

For the image-based classification experiments,
we include results using both the GP mean prediction and the NN prediction for both \our and the GP subset.
For the neural network training, we used the same set-up as in \citet{immer2021improving},
training the neural networok using Adam for $500$ epochs using a batch size of $512$.
The results reported in \cref{tbl:imagesuper} used a learning rate of $10^{-4}$
and we ran each experiment for $5$ seeds.
The MNIST experiments used a prior precision of $\delta=0.0008$
and the FMNIST experiments used a prior precision of $\delta=0.0013$.
The experiments with MNIST  used an MLP with two hidden layers of width 128 and Tanh activtions.
In the FMNIST experiments we followed \cite{immer2021improving} and used a CNN
architecture with two fully connected classification layers and Tanh activations.

In contrast to the GP subset implementation in \citet{immer2021improving},
we do not perform scaling of $\delta$ after the neural network has been trained.
Instead, we use the same prior precision that was used during training.
We performeed a grid search over the prior precision for the NN MAP using the validation NLPD.
We then used this prior precision for the GP predictive (NN) with $M=1024$,
GP predictive with $M=1024$, SFR (NN) and SFR.
We found the MNIST and FMNIST experiments to be numerically stable, even when not performing any clipping in computing the Hessian of the log likelihood, or adding jitter before Cholesky decomposition within the \our dual computation.
